---
title: 'The New Carpentries Lesson Template: {sandpaper}'
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  eval = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-", out.width = "100%")
```

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/sandpaper)](https://CRAN.R-project.org/package=sandpaper)
[![R build status](https://github.com/zkamvar/sandpaper/workflows/R-CMD-check/badge.svg)](https://github.com/zkamvar/sandpaper/actions)
[![Codecov test coverage](https://codecov.io/gh/zkamvar/sandpaper/branch/master/graph/badge.svg)](https://codecov.io/gh/zkamvar/sandpaper?branch=master)
<!-- badges: end -->

The {sandpaper} package was created by [The Carpentries] to re-imagine our method
of creating lesson websites for our workshops. This package will take a series 
of [Markdown] or [RMarkdown] files and generate a static website with the 
features and styling of The Carpentries lessons including customized layouts and
callout blocks. Much of the functionality in this package is inspired by
[Jenny Bryan's](https://jennybryan.org/) work with the [{usethis}] package.

--------------------------------------------------------------------------------

> Please note: {sandpaper} is very much a work in progress. At the moment, the
> only thing that is real is the README to serve as a roadmap for the future.
> Please [open an issue](https://github.com/zkamvar/sandpaper/issues/new) if you
> have any comments or suggestions!

--------------------------------------------------------------------------------

## Installation

{sandpaper} is not currently on CRAN, but it can be installed from github via
the [{remotes}] package:

``` r
# install.packages("remotes")
remotes::install_github("zkamvar/sandpaper", dep = TRUE)
```

Note that this will also install development versions of the following packages:

| package     | What it does | 
| ------------| ------------ |
| [{varnish}] | html, css, and javascript templates for The Carpentries (in progress) |
| [{usethis}] | Helper tools for R users that makes the experience of contribution easier |
| [{gh}] | interface to the github API |


## In-progress example

An in-progress example of this package can be found at https://github.com/zkamvar/miniature-octo-waffle

## Design

This package is designed to make the life of the lesson contributors and
maintainers easier by separating the tools needed to build the site from the 
user-defined content of the site itself. It will no longer rely on Jekyll or
any of the other [>450 static site generators](https://staticsitegenerators.net),
but instead rely on R, RStudio, and [{pkgdown}] to generate a site with the
following features:

- [ ] optional offline use
- [x] filename-agnostic episode arrangements
- [ ] clear definitions of package versions needed to build the lesson
- [ ] lesson versioning (e.g. I can navigate to
   https://swcarpentry.github.io/python-novice-gapminder for the current
   version and https://swcarpentry.github.io/python-novice-gapminder/2020-11
   for the release in 2020-11)
- [ ] seamless updates to the Carpentries' style
- [x] caching of rendered content for rapid deployment
- [ ] packaging of [{learnr}] materials
- [ ] validation of lesson structure
- [x] git aware, but does not require contributors to have git installed

### Rendering locally

In a repository generated via {sandpaper}, only the source is committed to avoid
issues surrounding out-of-date artefacts and directory structure confusion.

The website is generated in two steps:

1. markdown files from the source files are rendered containing a hash for the
   source file so that these need only be re-rendered when they change. 
2. html files are generated from the rendered markdown files and the CSS and JS
   sources in the [{varnish}] package for the preview.

To ensure there are no clashes between minor differences in the user setup, no
artifacts are committed to the main branch of the repository. Because of the
caching mechanism between the website and the rendered markdown files, 
long-running lessons can be updated and previewed quickly. 

### Rendering on continuous integration

Continuous integration will act as the single source-of-truth for how the
outputs of the lessons are rendered. For this, we want the resulting website
to be:

 - CI agnostic (but currently set up with GitHub)
 - easy to set up
 - auditable (e.g. I can see changes between the content of two commits)
 - versionable (e.g. I can instruct learners to go to `<WEBSITE>/1.1`. This 
   is inspired from the python documentation style)

To acheive this, there will be two branches created: `md-outputs` and `gh-pages`
that will inerit like so main -> `md-outputs` -> `gh-pages`. Because the build
time from main to `md-outputs` can be time intensive, this will default to
updating only files that were changed. 

 - `md-outputs`: this branch will contain the files and artifacts generated from
   rmarkdown in the vignettes directory of a thin package skeleton. 
 - `gh-pages`: this branch is generated via `md-outputs` and bundles the html,
   css, and js for the website. This will contain a single `index.html` file
   with several subfolders with different versions of the site. The `index.html`
   file will redirect to the `current/` directory, which contains the up-to-date
   site. 

> I believe that management of these can be done via `git worktree`, but I still
> need to dig into the mechanics of pkgdown further.

#### Scheduled builds

 - `gh-pages` website: Because we are designing the lessons to have content
   separated from the styling, we will set up the CI to generate the webpage
   from the pre-built sources on a weekly basis, which will check if there has
   been an update to the styles (which I have in the [{varnish}] package) and
   then rebuild the site without rebuilding the content. 
 - `md-outputs` branch: This will be rerun every month from scratch with the
   most recent version of R and R packages. If there is a change, a pull request
   can be generated to update the `renv.lock` file with a link to the changed
   markdown files in this branch. 


### Function syntax

The functions in {sandpaper} have the following prefixes:

 - `create_` will create/amend files or folders in your workspace
 - `build_` will build files from your source
 - `check_` validates either the elements of the lesson and/or episodes
 - `fetch_` will download files or resources from the internet
 - `reset_` removes files or information
 - `get_` will retrieve information from your source files as an R object
 - `set_` will update information in files.
 - `ci_` interacts with continous integration to build the website

Here is a working list of user-facing functions:

**Lesson and Episode Creation**

 - `create_lesson()` creates a lesson from scratch
 - `create_episode()` creates a new episode with the correct number prefix
 - `create_dataset()` creates a csv or text data set from an R object
 - `set_episodes()` arranges the episodes in a user-specified order

Accessors

 - `get_config()` reads the contents of `config.yaml` as a list
 - `get_episodes()` returns the episode filenames as a vector
 - `get_syllabus()` returns the syllabus with timings, titles, and questions

**Website Creation and Validation**

 - `check_lesson()` checks and validates the source files and lesson structure
 - `build_episode_md()` renders an individual file to markdown (internal use)
 - `build_episode_html()` renders a built markdown file to html (internal use)
 - `build_lesson()` builds the lesson into a static website
 - `build_portable_lesson()` builds the lesson into a portable static website
 - `fetch_lesson()` fetches the static website from the lesson repository

**Continuous Integration Utilities**

 - `ci_deploy()` builds and deploys the lesson on CI from the source files
 - `ci_deploy_site()` deploys the lesson on CI from pre-rendered markdown files
 - `ci_release()` builds and deploys the lesson on CI from the source files and adds a release tag

Cleanup

 - `reset_episodes()` removes the schedule from the config.yaml file
 - `reset_site()` clears the website and cache 

## Usage

There are five use-cases for {sandpaper}:

 1. Creating lessons
 2. Contributing to lessons
 3. Maintaining lessons
 4. Rendering a portable site
 5. Rendering a site with GitHub actions.

### Creating a lesson

To create a lesson with {sandpaper}, use the `create_lesson()` function:

```{r}
sandpaper::create_lesson("~/Desktop/r-intermediate-penguins")
```

This will create folder on your desktop called `r-intermediate-penguins` with
the following structure:

```
|-- .gitignore               # - Ignore everything in the site/ folder
|-- .github/                 # - Scripts used for continuous integration
|   `-- workflows/           #
|       |-- deploy-site.yaml # -   Build the source files on github pages
|       |-- build-md.yaml    # -   Build the markdown files on github pages
|       `-- cron.yaml        # -   reset package cache and test
|-- episodes/                # - PUT YOUR MARKDOWN FILES IN THIS FOLDER
|   |-- data/                # -   Data for your lesson goes here
|   |-- figures/             # -   All static figures and diagrams are here
|   |-- files/               # -   Additional files (e.g. handouts) 
|   `-- 00-introducition.Rmd # -   Lessons start with a two-digit number
|-- instructors/             # - Information for Instructors
|-- learners/                # - Information for Learners
|   `-- Setup.md             # -   setup instructions (REQUIRED)
|-- profiles/                # - Learner and/or Instructor Profiles
|-- site/                    # - This folder is where the rendered markdown files and static site will live
|   `-- README.md            # -   placeholder
|-- config.yaml              # - Use this to configure commonly used variables
|-- CONTRIBUTING.md          # - Carpentries Rules for Contributions (REQUIRED)
|-- CODE_OF_CONDUCT.md       # - Carpentries Code of Conduct (REQUIRED)
|-- LICENSE.md               # - Carpentries Licenses (REQUIRED)
`-- README.md                # - Introduces folks how to use this lesson and where they can find more information.
```

Once you have your site set up, you can add your RMarkdown files in the episodes
folder. By default, they will be built in alphabetical order, but you can use
the `set_episodes()` command to build the schedule in your `config.yaml` file:

```{r}
s <- sandpaper::get_episodes()
sandpaper::set_episodes(order = s, write = TRUE)
```

When you want to preview your site, use the following:

```{r}
sandpaper::build_lesson()
```

This will create the website structure inside of the the `site/` folder, render
the RMarkdown files to markdown (for inspection and quick rendering), render the
markdown files to HTML, and then enable a preview within your browser window.


### Contributing to a Lesson

To contribute to a lesson, you can either fork the lesson to your own repository
and clone it to your computer manually from GitHub, or you can use the {usethis}
package to automate it. For example, This is how you can create a copy of 
[**Programming With R**](http://swcarpentry.github.io/r-novice-inflammation/) to
your computer's Desktop.


```{r}
usethis::create_from_github(
  repo = "swcarpentry/r-novice-gapminder", 
  destdir = "~/Desktop/r-novice-gampinder",
  fork = TRUE
)
```

This will copy all of the source files to your computer and move you to the
directory.

Note that the rendered website will not be immediately available. To download
the site as it currently appears on the web, use:

```{r}
sandpaper::fetch_lesson(markdown = TRUE, site = TRUE)
```

This will download the site and the rendered markdown files into the `site/`
folder. To save bandwidth, you can choose to just download the markdown files
and artifacts by settin `site = FALSE`. Now, you can edit the Rmarkdown files
in `episodes/` and quickly render the site.

To upload changes to the lesson repository, you can use the follow


### Maintaining a Lesson

When you are maintaining a lesson, there is a high likelihood that you will
already have a copy on your machine. If not, follow the instructions in the
[contributing to a lesson](#contributing-to-a-lesson) section above.

The typical workflow will look like this:

1. open the sandpaper project in RStudio and make edits to files in the 
   `episodes/` folder
2. in the R console run the following 


```{r}
sandpaper::check_lesson() # validates the structure of the input files
sandpaper::build_lesson() # builds and validates lesson
```

### Rendering a portable site

To render a portable site, you can follow the instructions for [contributing
to a lesson](#contributing-to-a-lesson) or [maintaining a
lesson](#maintaining-a-lesson) to set up. Once you have the lesson set up, you
can use the following command:

```{r}
sandpaper::build_portable_lesson(version = "current")
```

This will render a fully portable lesson site as a zip file in the `site/`
folder. You can distribute this lesson to learners who do not have reliable 
internet access for use offline without sacrificing any of the styling. 

### Rendering with GitHub actions

Ultimately, there should be a minimal number of functions that handle this
situation because writing CI configuration files is maddening. The most
straightforward function is:

```{r}
sandpaper::ci_deploy(md_branch = "md-outputs", site_branch = "gh-pages")
```

This will render the markdown files in the `vignettes/` folder of the
"md-outputs" branch and then build the site from that folder and deploy it to
the "gh-pages" branch using the [{pkgdown}] machinery. 

Because css and js libraries may need updating before any lesson material does,
a step can be created just for rebuilding the site that uses:

```{r}
sandpaper::ci_deploy_site(branch = "gh-pages")
```

When a lesson is given a release, the current site folder needs to be duplicated
to a versioned folder and a tag needs to be added to the "md-outputs" branch:

```{r}
sandpaper::ci_release(tag = "0.1", md_branch = "md-outputs", site_branch = "gh-pages")
```


[The Carpentries]: https://carpentries.org
[Markdown]: https://daringfireball.net/projects/markdown/
[RMarkdown]: https://rmarkdown.rstudio.com/
[{learnr}]: https://rstudio.github.io/learnr/index.html
[{remotes}]: https://remotes.r-lib.org/
[{usethis}]: https://usethis.r-lib.org/
[{gh}]: https://gh.r-lib.org/
[{varnish}]: https://github.com/zkamvar/varnish#readme
[{gert}]: https://github.com/r-lib/gert#readme
[{testthat}]: https://github.com/r-lib/testthat#readme
[{pkgdown}]: https://github.com/r-lib/pkgdown#readme
