---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  eval = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# sandpaper

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/sandpaper)](https://CRAN.R-project.org/package=sandpaper)
<!-- badges: end -->

The {sandpaper} package was created by [The Carpentries] to re-imagine our method
of creating lesson websites for our workshops. This package will take a series 
of [Markdown] or [RMarkdown] files and generate a static website with the 
features and styling of The Carpentries lessons including customized layouts and
callout blocks. 

## Installation

{sandpaper} is not currently on CRAN, but it can be installed from github via
the {remotes} package:

``` r
# install.packages("remotes")
remotes::install_github("carpentries/sandpaper")
```

## Usage

There are three use-cases for {sandpaper}:

 1. Creating lessons
 2. Maintaining lessons
 3. Contributing to lessons
 4. Rendering a portable site

The functions in {sandpaper} have the following prefixes:

 - `create_` will create files or folders in your workspace
 - `build_` will build files from your source
 - `get_` will retrieve information from your source files as an R object
 
# Creating a lesson

To create a lesson with {sandpaper}, use the `create_lesson()` function:

```{r}
sandpaper::create_lesson("~/Desktop/r-intermediate-penguins")
```

This will create folder on your desktop called `r-intermediate-penguins` with
the following structure:


```
|-- .gitignore               # - Ignore everything in the site/ folder
|-- .github/                 # 
|   `-- workflows/           #
|       `-- workshop.yaml    # - Automatically build the source files on github pages
|-- episodes/                # - PUT YOUR MARKDOWN FILES IN THIS FOLDER
|   |-- data/                # - Data for your lesson goes here
|   |-- extras/              # - Supplemental lesson material goes here
|   |-- figures/             # - All static figures and diagrams are here
|   |-- files/               # - Additional files (e.g. handouts) 
|   `-- 00-introducition.Rmd # - Lessons start with a two-digit number
|-- site/                    # - This folder is where your static site will live 
|   `-- README.md            #
|-- config.yaml              # - Use this to configure commonly used variables
|-- CODE_OF_CONDUCT.md       # - Carpentries Code of Conduct (REQUIRED)
`-- README.md                # - Use this to tell folks how to contribute
```


[The Carpentries]: https://carpentries.org
[Markdown]: https://daringfireball.net/projects/markdown/
[RMarkdown]: https://rmarkdown.rstudio.com/
