---
title: "Building Lessons With A Package Cache"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building Lessons With A Package Cache}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE, message = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  error = TRUE,
  comment = "#|"
)
```

## Quickstart

If you are working on a lesson that has content with generated output, then this
guide is for you. This will get you started using a local package cache separate
from your default library for your lessons that ensure you can reproduce your
lesson locally. The {sandpaper} engine is mindful of lessons with generated
content in the following ways:

- *reliable setup*: the version of the lesson built on the carpentries website
  will be the same as what you build on your computer because the packages will
  be identical

- *environmentally friendly*: The lesson dependencies are NOT stored in your
  default R library and they will not alter your R environment.

- *transparent*: any additions or deletions to the cache will be recorded in
  the lockfile, which is tracked by git.

To get started, you need to *explicitly* give {sandpaper} permission to create
and use a cache with the function `use_package_cache()`

```{r setup}
library("sandpaper")
```

```{r use_package_cache, eval = FALSE}
use_package_cache()
```
```{r output, echo = FALSE}
msg <- readLines(system.file("resources/WELCOME", package = "renv"))
msg <- gsub("${RENV_PATHS_ROOT}", dQuote("/path/to/cache"), msg, fixed = TRUE)
choices <- sandpaper:::message_package_cache(msg)
cat(paste(c("1:", "2:"), choices), sep = "\n")
```

> NOTE: If you have already used {renv} previously, then you will not get this
> notification. Instead, you will get a notification that your consent has 
> already been provided.

When you select `1` for yes, your global package cache will be created by the
{renv} package and sandpaper will automatically detect and install any packages
(and updates) needed for your lesson. **Once you have set up your cache, it can
be used for *all* of your lessons going forward.**

For example, let's say you have a lesson that uses the {cowsay} and {curl} 
packages to bring a random cat fact at the end of the lesson:

````{.markdown}
::::::::::::::::::::::::::::::::::::: keypoints

- Use `.Rmd` files for lessons even if you don"t need to generate any code
- Run `sandpaper::check_lesson()` to identify any issues with your lesson
- Run `sandpaper::build_lesson()` to preview your lesson locally

::::::::::::::::::::::::::::::::::::::::::::::::

```{r catfact, message=FALSE, echo=FALSE}`r ""`
library(curl)
library(cowsay)
say("catfact", "cat")
```
````


When you build your lesson, you will see this kind of output:

```{r build-lesson-dummy, eval = FALSE}
build_lesson()
```

```{.output}
#| ℹ Consent to use package cache provided
#| → Searching for and installing available dependencies
#| * Discovering package dependencies ... Done!
#| * Copying packages into the library ... Done!
#| * Resolving missing dependencies  ... 
#| Retrieving "https://cran.rstudio.com/src/contrib/cowsay_0.8.0.tar.gz" ...
#|  OK [downloaded 564.5 Kb in 0.1 secs]
#| Retrieving "https://cran.rstudio.com/src/contrib/crayon_1.4.1.tar.gz" ...
#|  OK [downloaded 34.9 Kb in 0.1 secs]
#| Retrieving "https://cran.rstudio.com/src/contrib/fortunes_1.5-4.tar.gz" ...
#|  OK [downloaded 188.4 Kb in 0.1 secs]
#| Retrieving "https://cran.rstudio.com/src/contrib/rmsfact_0.0.3.tar.gz" ...
#|  OK [downloaded 10.5 Kb in 0.1 secs]
#| Installing crayon [1.4.1] ...
#|  OK [built from source]
#| Copying crayon [1.4.1] into the cache ...
#|  OK [copied to cache in 5.9 milliseconds]
#| Installing fortunes [1.5-4] ...
#|  OK [built from source]
#| Copying fortunes [1.5-4] into the cache ...
#|  OK [copied to cache in 5.6 milliseconds]
#| Installing rmsfact [0.0.3] ...
#|  OK [built from source]
#| Copying rmsfact [0.0.3] into the cache ...
#|  OK [copied to cache in 5.7 milliseconds]
#| Installing cowsay [0.8.0] ...
#|  OK [built from source]
#| Copying cowsay [0.8.0] into the cache ...
#|  OK [copied to cache in 7.5 milliseconds]
#| → Restoring any dependency versions
#| * The library is already synchronized with the lockfile.
#| → Recording changes in lockfile
#| The following package(s) will be updated in the lockfile:
#| 
#| # CRAN ===============================
#| - cowsay     [* -> 0.8.0]
#| - crayon     [* -> 1.4.1]
#| - curl       [* -> 4.3.2]
#| - fortunes   [* -> 1.5-4]
#| - rmsfact    [* -> 0.0.3]
#| 
#| * Lockfile written to "/path/to/lesson/renv/profiles/lesson-requirements/renv.lock".
#| ℹ Using package cache in /path/to/cache
#| 
#| [ truncated ]
```

The output will use the newly downloaded packages:

````{.markdown}
::::::::::::::::::::::::::::::::::::: keypoints

- Use `.Rmd` files for lessons even if you don"t need to generate any code
- Run `sandpaper::check_lesson()` to identify any issues with your lesson
- Run `sandpaper::build_lesson()` to preview your lesson locally

::::::::::::::::::::::::::::::::::::::::::::::::

```{.output}

 -------------- 
A cat can’t climb head first down a tree because every claw on a cat’s paw points the same way. To get down from a tree, a cat must back down.
 --------------
    \
      \
        \
            |\___/|
          ==) ^Y^ (==
            \  ^  /
             )=*=(
            /     \
            |     |
           /| | | |\
           \| | |_|/\
      jgs  //_// ___/
               \_)
  
```
````



### Temporarily turning off the cache

There are times when you might want to not use the cache for a lesson. To do 
this you can use `no_package_cache()` and your lesson will default to using
your local library. 

```{r build-lesson-dummy2, eval = FALSE}
no_package_cache()
sandpaper:::build_markdown(rebuild = TRUE)
```

```{.output}

#| ℹ Consent to use package cache not given. Using default library.
#| → use `use_package_cache()` to enable the package cache for reproducible builds.
#| → You can switch between using your cache and the default library with `options(sandpaper.use_renv = TRUE)` (`FALSE` for the default library)
#| 
#| [ truncated ]
```

The output will use packages if they are available, but throw errors if they
do not exist:

````{.markdown}
::::::::::::::::::::::::::::::::::::: keypoints

- Use `.Rmd` files for lessons even if you don"t need to generate any code
- Run `sandpaper::check_lesson()` to identify any issues with your lesson
- Run `sandpaper::build_lesson()` to preview your lesson locally

::::::::::::::::::::::::::::::::::::::::::::::::

```{.error}
Error in library(cowsay): there is no package called "cowsay"
```

```{.error}
Error in say("catfact", "cat"): could not find function "say"
```
````



When you want to turn it on again, you can use `use_package_cache()`

```{r, eval = FALSE}
use_package_cache()
```
```{.output}
#| ℹ Consent for renv provided---consent for package cache implied.
```

## Motivation

One of the biggest motivations for building the lesson infrastructure around
{sandpaper} is to be mindful of how we build lessons with generated content. For
the moment, these are explicitly R Markdown lessons, but in the future, it is
possible for us to expand to different engines *without changing your lesson's*
setup.

In the past, we've done this by including a script inside the lesson template
that would automatically scan the dependencies of your lesson and install them
to your computer. While convenient, there were drawbacks to this method in that
it would alter your personal R library if any of the packages update, which was
not good if you were using R for you own work and wanted to preserve the package
versions that you had.

The ‘renv’ package provides a very useful interface to bring one aspect of
reproducibility to R projects. Because people working on Carpentries lessons
are also working academics and will likely have projects on their computer
where the package versions are necessary for their work, it's important that
those environments are respected.

Our flavor of renv applies a package cache explicitly to the content of the
lesson, but does not impose itself as the default renv environment.

This provisioner will do the following steps:

 1. check for consent to use the package cache via ‘use_package_cache()’ and
    prompt for it if needed

 2. check if the profile has been created and create it if needed via
    ‘renv::init()’

 3. populate the cache with packages needed from the user's system and download
    any that are missing via ‘renv::hydrate()’. This includes all new packages
    that have been added to the lesson.

 4. If there is a lockfile already present, make sure the packages in the cache
    are aligned with the lockfile (downloading sources if needed) via
    ‘renv::restore()’.

 5. Record the state of the cache in a lockfile tracked by git.  This will
    include adding new packages and removing old packages. ‘renv::snapshot()’

When the lockfile changes, you will see it in git and have the power to either
commit or restore those changes.

