---
title: "Data and Flow from Source to Website"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data and Flow from Source to Website}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "##"
)
```

## Introduction

This is a vignette that is designed for R package developers who are looking to
understand how the data flows between the lesson source and the final website.
I am assuming that the person reading this has familiarity with R packaging and
R environments.

One of the design philosophies of The Workbench is that if a lesson author or
contributor would like to add any sort of metadata or modify a setting in any
given lesson, they can do so by editing the source of the lesson with no need to
modify their workflow.

Metadata is different from content because, while it is not directly related to
the content, it has extra information that helps the users of the site navigate
the content. For example, each episode page has three piece of metadata
embedded as a YAML list at the top of the source file that defines the title
along with estimated time for teaching and excercises.

```yaml
title: "Introduction"
teaching: 5
exercises: 5
```

In terms of the lesson itself, metadata that is related to the whole lesson is
stored in `config.yaml`. This YAML file is designed to be as flat as possible to
avoid common problems with writing YAML. Metadata here include things like the
title of the Lesson, the source page of the lesson, the time it was created, and
the lesson program it belongs to.

```yaml
title: "An Example Lesson"
carpentry: "incubator"
created: "2023-11-27"
life-cycle: "pre-alpha"
```

It also defines the order of the episodes and other content.

```yaml
episodes:
 - introduction.md
 - first-example.md
```


## Example

All of these metadata are collected and stored in memory before the lesson is
built (triggered during `validate_lesson()`), which are accessible via the
objects defined by the internal `sandpaper:::set_globals()`. Here I will set
up an example lesson that I will use to demonstrate these global variables. 
Please note that I will be using internal functions in this demonstration. These
internal functions are not guaranteed to be stable outside of the context of
{sandpaper}. 

```{r create-example}
library("sandpaper")
snd <- asNamespace("sandpaper")
# create a new lesson
lsn <- create_lesson(tempfile(), name = "An Example Lesson", 
  rstudio = TRUE, open = FALSE, rmd = FALSE)
# add a new episode
create_episode_md(title = "First Example", add = TRUE, path = lsn, open = FALSE)
```

Within {sandpaper}, there are environments that contain metadata related to the
whole lesson called `.resources`, `instructor_globals`, `learner_globals`, and
`this_metadata`. Before the lesson is validated, these values are empty:

```{r globals}
snd$.resources$get()
snd$instructor_globals$get()
snd$learner_globals$get()
snd$this_metadata$get()
```


When I run `validate_lesson()` the first time, all the metadata is collected and
parsed from the `config.yaml`, and the individual episodes and cached. 

```{r validate}
# first run is always longer than the second
system.time(validate_lesson(lsn))
system.time(validate_lesson(lsn))
snd$.resources$get()
snd$instructor_globals$get()
snd$learner_globals$get()
snd$this_metadata$get()
snd$get_path_site_yaml(lsn)
```



## Static Variables

### Build Options

### Metadata

## Caculated Variables

### Site-wide metadata

### Page specific data
