---
title: "Data and Flow from Source to Website"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data and Flow from Source to Website}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "##"
)
```

## Introduction

This is a vignette that is designed for R package developers who are looking to
understand how the data flows between the lesson source and the final website.
I am assuming that the person reading this has familiarity with 
[R packaging](https://r-pkgs.org/) and [R
environments](https://r-pkgs.org/data.html#sec-data-state).

One of the design philosophies of The Workbench is that if a lesson author or
contributor would like to add any sort of metadata or modify a setting in any
given lesson, they can do so by editing the source of the lesson with no need to
modify their workflow.

### Two Sources of Metadata

Metadata is different from content because, while it is not directly related to
the content, it has extra information that helps the users of the site navigate
the content. For example, each episode page has three piece of metadata
embedded as a YAML list at the top of the source file that defines the title
along with estimated time for teaching and excercises.

```yaml
title: "Introduction"
teaching: 5
exercises: 5
```

In terms of the lesson itself, metadata that is related to the whole lesson is
stored in `config.yaml`. This YAML file is designed to be as flat as possible to
avoid common problems with writing YAML. Metadata here include things like the
title of the Lesson, the source page of the lesson, the time it was created, and
the lesson program it belongs to.

```yaml
title: "An Example Lesson"
carpentry: "incubator"
created: "2023-11-27"
life-cycle: "pre-alpha"
```

It also defines optional parameters such as the order of the episodes and other content that can be used to customise how the lesson is built.

```yaml
episodes:
 - introduction.md
 - first-example.md

handout: true
```

The thing to know about this metdata and these variables is that they all get
passed to {varnish} and are used to control how the lesson website is built
along with the metadata.

### An introduction to `{varnish}`

In order to understand how to pass data from `config.yaml` to {varnish}, it is
important to first understand the paradigm of how {sandpaper} and {varnish} work
together to produce a lesson website. Behind the scenes, we build markdown with
{knitr}, render the raw HTML with Pandoc and then use {pkgdown} to insert the
HTML and metadata into a template (written in [the logicless templating
language, Mustache](https://mustache.github.io/mustache.5.html)) that can be
updated and modified independently of {sandpaper}. This template is called
{varnish}.

In the paradigm of {pkgdown}, the HTML template is split up into different
components that are rendered separately and then combined into a single layout
template. For example, here is the template that defines the layout:


````{r varnish-template-layout, echo = FALSE, results = "asis"}
writeLines("```html")
tplt <- system.file("pkgdown", "templates", "layout.html", package = "varnish")
writeLines(readLines(tplt))
writeLines("```")
````

You can see that it contains `{{{ head }}}`, `{{{ header }}}`, `{{{ navbar }}}`,
`{{{ content }}}` and `{{{ footer }}}`. These are all the results of the other
rendered templates. For example, here's the `head.html` template, which defines
the content that goes in the HTML `<head>` tag (defining titles, metadata, 
stylesheets, and JavaScript):


````{r varnish-template-head, echo = FALSE, results = "asis"}
writeLines("```html")
tplt <- system.file("pkgdown", "templates", "head.html", package = "varnish")
writeLines(readLines(tplt))
writeLines("```")
````


## Storage Function Factories

There are two types of storage [function
factories](https://adv-r.hadley.nz/function-factories.html) in {sandpaper}:
Lesson Store (`.lesson_store()`) and List Store (`.list_store()`). Both of these
return lists of functions that access their calling environments, which are
created when the package is loaded:

### List Store

The list store acts like a persistent list that has `get()` `set()`, `clear()`,
`copy()` and `update()` methods.

```{r list-store}
library("sandpaper")
snd <- asNamespace("sandpaper")
this_list <- snd$.list_store()
names(this_list)
class(this_list)

# to set a list, use a NULL key
this_list$set(key = NULL, list(
  A = list(first = 1),
  B = list(
    C = list(
      D = letters[1:4],
      E = sqrt(2),
      F = TRUE
    ),
    G = head(sleep)
  )
))

# get the list
this_list$get()

# copy a list so that you can preserve the original list
that_list <- this_list$copy()

# update list elements by adding to them (note: vectors will be replaced)
that_list$update(list(A = list(platform = "MY COMPUTER")))
that_list$get()

# set nested list elements
that_list$set(c("A", "platform"), "YOUR COMPUTER")

# note that modified copies do not modify the originals
that_list$get()[["A"]]
this_list$get()[["A"]]
```


### Lesson Store

This creates the object containing the `pegboard::Lesson` object and it is 
a special object because when a lesson is set with this object, it will
additionally set the _other_ global data (as shown below).

```{r lesson-store}
snd <- asNamespace("sandpaper")
some_lesson <- snd$.lesson_store()
names(some_lesson)
```

## Example

All of these metadata are collected and stored in memory before the lesson is
built (triggered during `validate_lesson()`), which are accessible via the
objects defined by the internal `sandpaper:::set_globals()`. Here I will set
up an example lesson that I will use to demonstrate these global variables. 
Please note that I will be using internal functions in this demonstration. These
internal functions are not guaranteed to be stable outside of the context of
{sandpaper}. Using the `asNamespace("sandpaper")` function allows me to create
a method for accessing the internal functions:

```{r create-namespace-example, message = FALSE}
library("sandpaper")
snd <- asNamespace("sandpaper")
```

```{r, create-example, message = FALSE}
# create a new lesson
lsn <- create_lesson(tempfile(), name = "An Example Lesson", 
  rstudio = TRUE, open = FALSE, rmd = FALSE)
# add a new episode
create_episode_md(title = "First Example", add = TRUE, path = lsn, open = FALSE)
```

Within {sandpaper}, there are environments that contain metadata related to the
whole lesson called `.store`, `.resources`, `instructor_globals`,
`learner_globals`, and `this_metadata`. Before the lesson is validated, these
values are empty:

```{r globals}
class(snd$.store)
print(snd$.store$get())
class(snd$.resources)
snd$.resources$get()
class(snd$instructor_globals)
snd$instructor_globals$get()
class(snd$learner_globals)
snd$learner_globals$get()
class(snd$this_metadata)
snd$this_metadata$get()
```

These will contain the following information:

 - `.store` a `pegboard::Lesson` object that contains the XML representation of
   the parsed Markdown source files
 - `.resources` a list of the availabe files used to build the lesson in the
   order specified by `config.yaml`
 - `instructor_globals` pre-computed global data that includes the sidebar
   template, the syllabus, the dropdown menus, and information about the
   packages used to build the lesson
 - `learner_globals` same as `instructor_globals`, but specifically for learner
   view
 - `this_metadata` the real and computed metadata associated with the lesson
   along with the JSON-LD template for generating the metadata in the footer of
   the lesson.

When I run `validate_lesson()` the first time, all the metadata is collected and
parsed from the `config.yaml`, and the individual episodes and cached.

```{r validate}
# first run is always longer than the second
system.time(validate_lesson(lsn))
system.time(validate_lesson(lsn))
```

The `validate_lesson()` call will pull from the cache in `.store` if it exists
and is valid (which means that nothing in git has changed). If it's not valid or
does not exist, then all of the global storage is initialised in this general
cascade:

```{r vl-tree, echo = FALSE}
funs <- c(
  vl = "validate_lesson()", 
  tl = "this_lesson()", 
  sv = ".store$valid()",
  ggd = "gert::git_diff()",
  ggs = "gert::git_status()", 
  ggl = "gert::git_log()", 
  pl = "pegboard::Lesson$new()",
  ss = ".store$set()", 
  sg = "set_globals()",
  srl = "set_resource_list()",
  res = ".resources$set()",
  grl = "get_resource_list()",
  im = "initialise_metadata()",
  gc = "get_config()",
  tm = "template_metadata()",
  tms = "this_metadata$set()",
  cs = "create_sidebar()",
  crd = "create_resources_dropdown()",
  lgs = "learner_globals$set()",
  igs = "instructor_globals$set()"
)
labels <- funs
to_lab <- c("ss", "res", "tms", "lgs", "igs")
labels[to_lab] <- cli::col_cyan(cli::style_bold(labels[to_lab]))
vltree <- data.frame(
  fn = funs,
  calls = I(list(
    vl = funs["tl"], 
    tl = list(funs["sv"], funs["pl"], funs["ss"]), 
    sv = list(funs["ggd"], funs["ggs"], funs["ggl"]),
    ggd = character(0),
    ggs = character(0),
    ggl = character(0), 
    pl = character(0),
    ss = list(funs["sg"]), 
    sg = list(funs["im"], funs["srl"], funs["cs"], funs["crd"], funs["lgs"],
      funs["igs"]), 
    srl = list(funs["grl"], funs["res"]),
    res = character(0),
    grl = list(funs["gc"]),
    im = list(funs["gc"], funs["tm"], funs["tms"]),
    gc = character(0),
    tm = character(0),
    tms = character(0),
    cs = character(0),
    crd = character(0),
    lgs = character(0),
    igs = character(0)
    )
  ),
  labels = labels
)
cli::tree(vltree)
```

Now when we these variables are called, you can see the information stored in
them:

### Lesson Storage

```{r store-get}
class(snd$.store)
print(snd$.store$get())
```

```{r resources-get}
snd$.resources$get()
snd$instructor_globals$get()
snd$learner_globals$get()
snd$this_metadata$get()
```


### pkgdown metadata

Another source of metadata is that created for the pkgdown in a file called
`site/_pkgdown.yaml`


```{r yaml, results="asis", echo = FALSE, comment = ""}
writeLines("```yaml")
writeLines(readLines(fs::path(lsn, "site", "_pkgdown.yaml")))
writeLines("```")
```




## Static Variables

### Build Options

### Metadata

## Caculated Variables

### Site-wide metadata

### Page specific data
